Effects of Different Pain Relievers and Dosages on Cognitive Retention

### 2025-04-02 paper/paper.qmd
References fixed
Format changes

---
title: "Effects of Different Pain Relievers and Dosages on Cognitive Retention"
subtitle: Results
author: 
  - Navya Hooda
  - Mohammed Yusuf Shaikh
  - Shivank Goel
  - Vanshika Vanshika
  - Jena Shah
date: today
date-format: long
abstract: Painkillers are commonly used to manage pain, and their effects are often associated with  physical relief. While they provide immediate physical comfort, their impact on cognitive performance is often overlooked. This study explores how different painkillers and dosages influence cognitive function. Using simulated testing from a virtual population, participants were randomly assigned a treatment (drug type and dosage) while controlling for variables such as age. Our findings indicate no significant relationship between drug type, dosage, and overall cognitive performance post-treatment. However, we observed some correlation between certain drug-dosage combinations and memory improvement, suggesting a potential area for further research. While the results were not statistically significant, this study contributes to a deeper understanding of the cognitive effects of painkillers and highlights the need for further investigation into their potential impact on memory and cognition.
number-sections: true
format: pdf
bibliography: references.bib
toc: true
---

@@

No auto-complete tools such as co-pilot were used in the course of this project, however, Language Learning Model ChatGPT was used while writing this paper. It was used for the purpose of code debugging, understanding models, and knowledge of certain topics, which we were not aware of. The chat with the AI bot is also attached as a reference under @sec-appendix

@@

Pain can make everyday tasks harder, particularly when mental focus and clarity are required. Study has shown that individuals with pain tend to have cognitive dysfunction symptoms (Hu, 2010). Common pain relievers like Aspirin, Paracetamol, and Tramadol are often used to reduce physical discomfort, but their impact on cognitive functions are not well understood. Memory retention is a process that requires mental effort – which becomes more difficult when we're in pain. 
This study investigates how different pain relievers (Aspirin, Paracetamol, and Tramadol) at standard dosages affect memory performance. We will also look at how the interaction between drug type and dosage level influence memory function. Since the ability to recall information is a basis for optimized human function, understanding how pain relief might impact cognitive performance is important for managing pain without affecting productivity.

## Research Questions
1. *How does the type of pain reliever (Aspirin 500 mg, Paracetamol 500 mg, Tramadol 50 mg, and Placebo) affect memory game performance?*
- Null Hypothesis: All pain relievers at normal dosages act the same for cognitive retention game performances.
- Alternative Hypothesis: There exists at least one pair of distinct pain relievers at normal dosages which act differently for cognitive retention game performances.

2. *Is there an interaction effect between the type of pain reliever and dosage level on the improvement in memory game performance?*
- Null Hypothesis: There is no interaction effect between the type of pain reliever and dosage level on the improvement in memory task performance.
- Alternative Hypothesis: There is an interaction effect between at least one of the types of pain reliever and dosage level on the improvement in memory task performance.

@@

## Study Design and Treatments   
We selected three brands of painkillers: Aspirin, Paracetamol, and Tramadol. Additionally, we considered a placebo treatment. As a result, this study involves 2 factors: pain reliever type (4 levels - Aspirin, Paracetamol, Tramadol, Placebo) and dosage (2 levels - Low, High). Crossing the two factors provides us with a total of *8 treatment groups*, with **30 participants** assigned to each. The groups are as follows:

# Create a dataframe for the table
study_table <- data.frame(
  Dosage = c("Low", "High"),
  Aspirin = c("500 mg", "1000 mg"),
  Paracetamol = c("500 mg", "1000 mg"),
  Tramadol = c("50 mg", "100 mg"),
  Placebo = c("Low", "High")
)

# Render the table
kable(study_table, caption = "Treatment Groups",
      align = "c") %>%
  kable_styling(full_width = FALSE, position = "center")
```

@@

### Nuisance Variable: Age  
To reduce bias, we consider **controlling** the experiment by limiting participant ages to be strictly 18+. Hence in this study, **age** acts as a **nuisance variable**.
We also used **blocking** for age by ensuring participants were categorized into three even groups to prevent bias created through age. @tbl-summary-5 shows the breakdown of the groups.

## Quantitative and Qualitative Variables 
In our study the quantitative and qualitative variables are as follows:

@@

@tbl-summary-2 shows the distribution of participants across drug types. Aspirin, Paracetamol, and Tramadol, and Placebo each have 60 participants (25\%), ensuring balance.

@tbl-summary-3 shows the dosage levels assigned to participants. There is an equal split between high-dose (120 participants, 50\%) and low-dose (120 participants, 50\%) conditions.

\newpage
# Plots

@@

```
```{r}
#| label: fig-combined
#| echo: false
#| warning: false
#| message: false
#| fig.width: 12
#| fig.height: 6
#| fig-cap: "Interaction Plot and Boxplot: Memory Game Scores by Drug Type and Dosage Level"

@@

- $Y_{ijk}$: Memory score for the $k$-th participant receiving the $i$-th drug and $j$-th dosage level  
- $\mu$: Overall mean  
- $\alpha_i$: Effect of the $i$-th drug ((Aspirin, Tramadol, Paracetamol, Placebo))  
- $\beta_j$: Effect of the $j$-th dosage level (High Dosage, Low dosage) 
- $(\alpha\beta)_{ij}$: Interaction effect between drug and dosage  
- $\epsilon_{ijk} \sim N(0, \sigma^2)$ : Random error

@@

- $\beta_0$: Intercept (Aspirin at High dosage — the reference group)  
- $\beta_1, \beta_2, \beta_3$: Main effects of Paracetamol, Placebo, and Tramadol relative to Aspirin  
- $\beta_4$: Effect of Low dosage relative to High  
- $\beta_5, \beta_6, \beta_7$: Interaction effects for each drug under Low dosage  
- $\epsilon_i$: Residual error for the $i$-th participant

@@

## need to work on it
"The two-way ANOVA factorial design involves the linear regression model followed by the ANOVA model. The model uses the reference (baseline) group which is Aspirin (Drug) & High (Dosage level). The regression model from the @tbl-model-result shows the intercept(beta_0) which represents the predicted mean of memory game score for participants receiving high dosage of aspirin is 62.20 and the p-value = 0.00000 indicates the intercept is statistically significant. The variance explained by the model is R^2 = 0.03983, approximately 4% of the response variable. And the p-value = 0.2167 of the model can be interpreted as not statistically significant"

# Discussion
## Interpretation of Results  
The two-way ANOVA analysis revealed no statistically significant main effects of drug type (p = 0.25) or dosage level (p = 0.26) on memory game scores. Similarly, the interaction effect between drug type and dosage level was not significant (p = 0.24). These results suggest that neither the type of pain reliever (Aspirin, Paracetamol, Tramadol, or Placebo) nor its dosage (low vs. high) had a meaningful impact on memory game scores, which is representative of cognitive improvement in this study. The lack of significance persisted even after accounting for age as a confounding variable, which was controlled through participant stratification into three balanced groups (18–34, 35–50, and 50+ years).

# Limitations 
Our study is overall focused on memory, which may have links to one’s education and IQ level as they are all related to mental function and cognition. In the future, we could test everyone for IQ scores and their highest level of education – these two nuisance variables can be taken care of by using the four principles of experimental design. Additionally, we used 30 observations for each treatment group which may not be large enough. We can increase this number for future studies. To add on, not all participants were in pain, so the efficacy of the improvement of pain killers may be biased and using a virtual population may not ensure whether this aspect could be captured. Lastly, cognitive retention was measured immediately post-treatment – the long term impacts are not yet explored. To improve our study, we can try to  consider individuals with real pain symptoms before the study to ensure a reduction of bias, and provide a realistic test setting. Additionally, carrying the experiment out in the long-term would help with understanding the effects of painkillers and dosages on memory.

# Conclusion
This study found no conclusive evidence that pain relievers or their dosages significantly affect cognitive retention. However, the exploratory trends observed, particularly with Paracetamol and Tramadol, highlight the need for larger-scale studies to investigate subtle or context-dependent effects. Future research could incorporate longitudinal designs, broader age-specific analyses, and alternative cognitive metrics to deepen understanding of painkillers’ cognitive implications.

>>>>

---
title: "Effects of Different Pain Relievers and Dosages on Cognitive Retention"
author: 
  - Navya Hooda
  - Mohammed Yusuf Shaikh
  - Shivank Goel
  - Vanshika Vanshika
  - Jena Shah
date: today
date-format: long
abstract: Painkillers are commonly used to manage pain, and their effects are often associated with  physical relief. While they provide immediate physical comfort, their impact on cognitive performance is often overlooked. This study explores how different painkillers and dosages influence cognitive function. Using simulated testing from a virtual population, participants were randomly assigned a treatment (drug type and dosage) while controlling for variables such as age. Our findings indicate no significant relationship between drug type, dosage, and overall cognitive performance post-treatment. However, we observed some correlation between certain drug-dosage combinations and memory improvement, suggesting a potential area for further research. While the results were not statistically significant, this study contributes to a deeper understanding of the cognitive effects of painkillers and highlights the need for further investigation into their potential impact on memory and cognition.
number-sections: true
format: pdf
bibliography: references.bib
toc: true
---

@@

No auto-complete tools such as co-pilot were used in the course of this project, however, the Language Learning Model 'ChatGPT' was used while writing this paper.
It was used for the purpose of code debugging, understanding models, and knowledge of certain topics, which we were not aware of.
The chat with the AI bot is also attached as a reference under @sec-appendix.

@@

Pain can make everyday tasks harder, particularly when mental focus and clarity are required.
Study has shown that individuals with pain tend to have cognitive dysfunction symptoms [@Hu].
Common pain relievers like Aspirin, Paracetamol, and Tramadol are often used to reduce physical discomfort, but their impact on cognitive functions are not well understood.
Memory retention is a process that requires mental effort – which becomes more difficult when we're in pain.
This study investigates how different pain relievers (Aspirin, Paracetamol, and Tramadol) at standard dosages affect memory performance.
We will also look at how the interaction between drug type and dosage level influence memory function.
Since the ability to recall information is a basis for optimized human function, understanding how pain relief might impact cognitive performance is important for managing pain without affecting productivity.

## Research Questions

Research Question 1.
*How does the type of pain reliever (Aspirin 500 mg, Paracetamol 500 mg, Tramadol 50 mg, and Placebo) affect memory game performance?*

-   Null Hypothesis: All pain relievers at normal dosages act the same for cognitive retention game performances.
-   Alternative Hypothesis: There exists at least one pair of distinct pain relievers at normal dosages which act differently for cognitive retention game performances.

Research Question 2.
*Is there an interaction effect between the type of pain reliever and dosage level on the improvement in memory game performance?*

-   Null Hypothesis: There is no interaction effect between the type of pain reliever and dosage level on the improvement in memory task performance.
-   Alternative Hypothesis: There is an interaction effect between at least one of the types of pain reliever and dosage level on the improvement in memory task performance.

@@

## Study Design and Treatments

We selected three brands of painkillers: Aspirin, Paracetamol, and Tramadol.
Additionally, we considered a placebo treatment.
As a result, this study involves 2 factors: pain reliever type (4 levels - Aspirin, Paracetamol, Tramadol, Placebo) and dosage (2 levels - Low, High).
Crossing the two factors provides us with a total of *8 treatment groups*, with **30 participants** assigned to each.
The total crossed treatment groups are shown in @tbl-total-groups.

@@

```{r}
#| label: tbl-total-groups
#| tbl-cap: "Treatment Groups"
#| echo: false
#| warning: false
#| message: false

# Create a dataframe for the table
study_table <- data.frame(
  Dosage = c("Low", "High"),
  Aspirin = c("Aspirin 500 mg", "Aspirin 1000 mg"),
  Paracetamol = c("Paracetamol 500 mg", "Paracetamol 1000 mg"),
  Tramadol = c("Tramadol 50 mg", "Tramadol 100 mg"),
  Placebo = c("Placebo 1 Tablet", "Placebo 2 Tablets")
)

# Render the table
kable(study_table, align = "c") %>%
  add_header_above(c(" " = 1, "Drug Type" = 4)) %>%
  kable_styling(full_width = FALSE, position = "center")
```

### Nuisance Variable: Age

To reduce bias, we consider **controlling** the experiment by limiting participant ages to be strictly 18+.
Hence in this study, **age** acts as a **nuisance variable**.
We also used **blocking** for age by ensuring participants were categorized into three even groups to prevent bias created through age.
@tbl-summary-5 shows the breakdown of the groups.

## Quantitative and Qualitative Variables

In our study the quantitative and qualitative variables are as follows:

@@

@tbl-summary-2 shows the distribution of participants across drug types.
Aspirin, Paracetamol, and Tramadol, and Placebo each have 60 participants (25%), ensuring balance.

@tbl-summary-3 shows the dosage levels assigned to participants.
There is an equal split between high-dose (120 participants, 50%) and low-dose (120 participants, 50%) conditions.

# Plots

@@

```{r}
#| label: fig-combined
#| echo: false
#| warning: false
#| message: false
#| fig.width: 12
#| fig.height: 6
#| fig-cap: "Interaction Plot and Boxplot: Memory Game Scores by Drug Type and Dosage Level"

@@

-   $Y_{ijk}$: Memory score for the $k$-th participant receiving the $i$-th drug and $j$-th dosage level\
-   $\mu$: Overall mean\
-   $\alpha_i$: Effect of the $i$-th drug ((Aspirin, Tramadol, Paracetamol, Placebo))\
-   $\beta_j$: Effect of the $j$-th dosage level (High Dosage, Low dosage)
-   $(\alpha\beta)_{ij}$: Interaction effect between drug and dosage\
-   $\epsilon_{ijk} \sim N(0, \sigma^2)$ : Random error

@@

-   $\beta_0$: Intercept (Aspirin at High dosage — the reference group)\
-   $\beta_1, \beta_2, \beta_3$: Main effects of Paracetamol, Placebo, and Tramadol relative to Aspirin\
-   $\beta_4$: Effect of Low dosage relative to High\
-   $\beta_5, \beta_6, \beta_7$: Interaction effects for each drug under Low dosage\
-   $\epsilon_i$: Residual error for the $i$-th participant

@@

## need to work on it

"The two-way ANOVA factorial design involves the linear regression model followed by the ANOVA model. The model uses the reference (baseline) group which is Aspirin (Drug) & High (Dosage level). The regression model from the @tbl-model-result shows the intercept(beta_0) which represents the predicted mean of memory game score for participants receiving high dosage of aspirin is 62.20 and the p-value = 0.00000 indicates the intercept is statistically significant. The variance explained by the model is R\^2 = 0.03983, approximately 4% of the response variable. And the p-value = 0.2167 of the model can be interpreted as not statistically significant"

# Discussion

## Interpretation of Results

The two-way ANOVA analysis revealed no statistically significant main effects of drug type (p = 0.25) or dosage level (p = 0.26) on memory game scores.
Similarly, the interaction effect between drug type and dosage level was not significant (p = 0.24).
These results suggest that neither the type of pain reliever (Aspirin, Paracetamol, Tramadol, or Placebo) nor its dosage (low vs. high) had a meaningful impact on memory game scores, which is representative of cognitive improvement in this study.
The lack of significance persisted even after accounting for age as a confounding variable, which was controlled through participant stratification into three balanced groups (18–34, 35–50, and 50+ years).

# Limitations

Our study is overall focused on memory, which may have links to one’s education and IQ level as they are all related to mental function and cognition.
In the future, we could test everyone for IQ scores and their highest level of education – these two nuisance variables can be taken care of by using the four principles of experimental design.
Additionally, we used 30 observations for each treatment group which may not be large enough.
We can increase this number for future studies.
To add on, not all participants were in pain, so the efficacy of the improvement of pain killers may be biased and using a virtual population may not ensure whether this aspect could be captured.
Lastly, cognitive retention was measured immediately post-treatment – the long term impacts are not yet explored.
To improve our study, we can try to consider individuals with real pain symptoms before the study to ensure a reduction of bias, and provide a realistic test setting.
Additionally, carrying the experiment out in the long-term would help with understanding the effects of painkillers and dosages on memory.

# Conclusion

This study found no conclusive evidence of pain relievers or their dosages significantly affecting cognitive retention.
However, the exploratory trends observed, particularly with Paracetamol and Tramadol, highlight the need for larger-scale studies to investigate subtle or context-dependent effects.
Future research could incorporate longitudinal designs, broader age-specific analyses, and alternative cognitive metrics to deepen understanding of painkillers’ cognitive implications.
